---
title: "Full 18s Community composition"
output: html_notebook
---

```{r libs, include=F}
library(tidyverse)
library(here)
library(GUniFrac)
library(ape)
library(ggtree)
library(ggrepel)
```

```{r input, include=F}
tree_path <- here("intermediate_output/tree/full_18s/all_18s_aligns.fasta.treefile")
all_18s_tree <- read.tree(tree_path)

ra_path <- here("data/clean/clean_18s_ra.csv")
ra <- read_csv(ra_path)

assignment_path <- here("data/clean/clean_silva_18s_assignments.csv")
assign <- read_csv(assignment_path)

field_path <- here("data/clean/field_clean.csv")
field <- read_csv(field_path)

alias_path <- here("data/sample_aliases.csv")
alias <- read_csv(alias_path)
```


```{r to_wide, include=F}
wide_ra <- ra %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  filter(asv_id %in% all_18s_tree$tip.label) %>% # Unifrac needs relative abundance table to match the tips on the tree
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")
head(wide_ra)
```



[Some background on Eukaryotic phylogeny](http://tolweb.org/Eukaryotes/3)
```{r root_tree, include=F}
# rooting by bacteria
root_id <- assign %>% 
  filter(kingdom=="Bacteria") %>% 
  head(1) %>% 
  pull(asv_id)

all_18s_tree <- all_18s_tree %>% root(outgroup = root, resolve.root=T)
```


```{r unifrac, include=F}
set.seed(123)
unifrac_out <- wide_ra %>% 
  GUniFrac::GUniFrac(all_18s_tree, alpha=c(0, 0.5, 1)) # ignore error message, we removed some otus from our tree, they should be automatically discluded

# select the weighted UniFrac matrix
weighted_unifrac_distm <- unifrac_out$unifracs[, , "d_1"]
dim(weighted_unifrac_distm)
```


```{r nmds_stress, include=F}
mono <- metaMDS(weighted_unifrac_distm)
mono
```


```{r unifrac_nmds, echo=F}
unifrac_plot_data <- mono %>%
  scores() %>%
  as_tibble(rownames = "sample_id") %>%
  left_join(field, by="sample_id")

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4, alpha=0.7) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  labs(color="Elevation (m)") +
  geom_text_repel(aes(label=sample_id)) +
  ggtitle("UniFrac NMDS of 18S amplicon data (only 2019 MiSeq run)")
  # geom_text(data = unifrac_plot_data %>%   # add sample ID labels
  #                   left_join(alias, by="sample_id"),
  #                   aes(label = sample_id), size=3,
  #                   position = position_nudge(y = -0.025))  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

  - Seymour samples clustered somewhat, except for sey18.25, sey18.63o
    - sey25 = brevispina, big biofilms
  - Hollyburn, 42b close to sey, but 21 and 47 were closer to high elevation samples
  - Sky12 closer to high elevation samples despite being from below treeline
  - Sax1, gar1, bre1, nes3 outliers from other high elevation samples

Check out sey18.25
```{r, echo=F}
ra %>% 
  filter(sample_id == "sey18.25") %>% 
  arrange(-rel_abund) %>% 
  left_join(assign, by="asv_id")
```


```{r set_plot, echo=F}
p <- ra %>% 
  left_join(assign, by="asv_id") %>% 
  group_by(phylum, sample_id) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  ungroup() %>% 
  left_join(field, by="sample_id") %>% 
  select(sample_id, phylum, rel_abund, date, elev_m) %>% 
  arrange(sample_id) %>% 
  filter(rel_abund >= 0.01) # only show phyla with >= 1% rel abund

# ggplot(p, aes(x=phylum, y=sample_id %>% fct_reorder(elev_m))) +
#   geom_point(aes(color = phylum, size = rel_abund), alpha=0.7) +
#   scale_size(range = c(1, 10)) +
#   ylab("Sample ID") +
#   xlab("OTU") +
#   labs(fill = "OTU", alpha="Relative abundance", title = "Relative abundance plot of 18S phyla ") +
#   scale_x_discrete(position = "bottom") +
#   theme(axis.text.x = element_text(hjust=1,angle=45)) +
#   scale_colour_discrete(guide = FALSE) # turn off legend
```


```{r facet_bar, echo=F, fig.width=10}
cust_theme <- theme(legend.position = "none",
                    panel.grid.major.x = element_blank(), # remove vertical lines in background
                    panel.grid.minor.x = element_blank(),
                    axis.line=element_blank(), # remove x axis
                    axis.title.x=element_blank(), 
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
                    strip.background = element_blank()) # remove grey boxes around facet labs

# order algae, fungi, protists, other
phylum_ord <- c("Chlorophyta_ph", "Ochrophyta","Basidiomycota", "Ascomycota", "Chytridiomycota", "Ciliophora","Cercozoa","Protalveolata", "Phragmoplastophyta")

ggplot(p, aes(x=sample_id %>% fct_reorder(elev_m), y=rel_abund)) +
  geom_bar(aes(fill=phylum), stat="identity") + # classic stacked relative abundance plot
  coord_flip() +
  facet_grid(~phylum %>% fct_relevel(phylum_ord), switch = "x") +
  labs(title = "Relative abundance plot of 18S phyla", x="Sample ID (from lowest to highest elevation)") +
  cust_theme
```

```{r, echo=FALSE}
view_phylum_table <- function(phylum){
  phylum <- enquo(phylum)
  ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(phylum == !!phylum) %>% 
  arrange(-rel_abund)
}
```

<br>
<br>


### Chlorophyta

####Trebouxiophyceae
```{r, echo=F}
ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(class == "Trebouxiophyceae") %>% 
  group_by(sample_id) %>% 
  summarise(rel_abund=sum(rel_abund)) %>% 
  arrange(-rel_abund)
```
  - sax18.01 outlier, much higher Trebouxiophyceae
    - could explain why an outlier in *rbc*L data?
    
####Chlorophyceae
```{r, echo=F}
ra %>% 
  left_join(assign, by="asv_id") %>% 
  filter(class == "Chlorophyceae") %>% 
  group_by(sample_id) %>% 
  summarise(rel_abund=sum(rel_abund)) %>% 
  arrange(rel_abund)
```
  - lowest Chlorophyceae is sey18.25, with 18%
  - highest " " is gar18.01, with 80%
  
### Ochrophyta
```{r, echo=F}
view_phylum_table("Ochrophyta")
```

  - higher abundance than exected
  - Xanthophyceae at pan18.01, otherwise mostly *Hydrurus*
  - Chrysophyceae, Leya found this family causing green snow
  - seen in microscopy
  - < 1% rel abundance
  - could be wind deposited or grow in snow
    - known to grow in cold, clear waters

### Basidiomycota
```{r, echo=F}
view_phylum_table("Basidiomycota")
```

  - sey7, sey25, hol21, pan.ro, wed2 have highest fungi
    - pan.ro:
      - once its off the snow, fungi jump in
      - high unknown Euks
    - 
  - Kriegeriales yeast
  - asv7 blast matches a bunch of fungi 100%
  
### Ascomycota
```{r, echo=F}
view_phylum_table("Ascomycota")
```
  - unknown Cordycipitaceae, a family of parasitic fungi
  - ASV 4 BLAST 100% *Fusarium*, widely distributed in soil, saprophyta assoc w plants (although Wiki places in a different family, Nectriaceae)

### Chytridiomycota
```{r, echo=F}
view_phylum_table("Chytridiomycota")
```

### Ciliophora:
```{r, echo=F}
view_phylum_table("Ciliophora")
```
  - SILVA assigns most to *Stokesia*, but best BLAST match for asv5 is *Frontonia* 
  - highest in hol18.42b (depth)
      - and nes3, green snow sample

### Cercozoa
```{r, echo=F}
view_phylum_table("Cercozoa")
```
  - Glissomonads
    - ASV17 BLAST best strain match, 96% *Orciraptor agilis*, a parasite of algae

### Arthropods
```{r, echo=F}
view_phylum_table("Arthropoda")
```

  - mostly at higher elev red snow sites, and hol47 (which LQ collected, dirty red snow near stream)
  - hol47 arthropods are all Collembola, similar to other sites

### Rotifera
```{r, echo=F}
view_phylum_table("Rotifera")
```
  - sey18.25: highest rotifers, matches photos. 
    - could explain why its an outlier in NMDS above
    - 










# Check differences between surface and depth samples:
  - 


# Check differences in high vs low diversity
  - 