---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
csl: article.csl
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m). Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome. 

  

---

# Introduction {-}

Placeholder


### *Keywords*: snow algae rbcL 18S amplicon microbiome alpine community
### *Run title*: Elevational patterns in snow algae communities
### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology
##### pagebreak
### Morphologically diverse snow algae blooms across a variety of habitats
### Diverse *rbc*L reveals species-level snow algae taxonomy
### Snow algae community composition varies with date and elevation
### Green snow algae turns orange at the surface
##### pagebreak
### *Supplementary information*.
### *Competing financial interests.* The authors declare that the research was conducted in the absence of any
##### pagebreak

<!--chapter:end:Casey1.Rmd-->

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  bookdown::word_document2()
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
cls: templates/article.cls
bibliography: casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m). Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# header 1
## header2

```{r}
plot(pressure)
```


<!--chapter:end:manuscript.Rmd-->

